Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/examples.tex
add a remark pointing to the supplemental material
various fixes

Because splitting \texttt{exp} does not eliminate its conditionals, partial
evaluation is more useful in this case. (Notice, however, that partially
evaluating \texttt{exp2} on a \texttt{binnat} is essentially the same as as
partially evaluating \texttt{exp} on the corresponding \rmint.) Nevertheless,
splitting \texttt{exp} still demonstrates how our algorithm finds interesting
data structures latent in the structure of recursive functions.

@@

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}

>>>>

Because splitting \texttt{exp} does not eliminate its conditionals, partial
evaluation is more useful in this case. (Notice, however, that partially
evaluating \texttt{exp2} on a \texttt{binnat} is essentially the same as 
partially evaluating \texttt{exp} on the corresponding \rmint.) Nevertheless,
splitting \texttt{exp} still demonstrates how our algorithm finds interesting
data structures latent in the structure of recursive functions.

@@

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}
%
In the appendix provided as supplemental material, we extend this example by
splitting \texttt{map (qss l)} over an arbitrary list of ranks \texttt{ks}. We also
provide an example of a program whose boundary data structure is a trie.

### 2015-07-09 paper/overview.tex
various fixes

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  Assuming that the input is uniformly
randomly ordered (which can be achieved by prepermuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since since
\texttt{s} is a list, leading to no improvement in efficiency.

>>>>

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

### 2015-07-09 paper/splitting.tex
various fixes

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
They are simply recursive on structure of the term, 
implying that the splitting algorithm runs in linear time.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on it's first stage components.
Analogously, this must hold of the output of splitting, 
which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.

>>>>

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
They are simply recursive on the structure of the term, 
implying that the splitting algorithm runs in linear time.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on its first stage components.
Analogously, this must hold of the output of splitting, 
which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.
